Quasi-universal acceptance of basic science in the US

### 2025-07-07 paper.Rmd
second round  revisions update preprint

  Substantial minorities of the population report a low degree of trust in science, or endorse conspiracy theories that violate basic scientific knowledge. This might indicate a wholesale rejection of science. In four studies, we asked 782 US participants questions about trust in science, conspiracy beliefs, and basic science (e.g. the relative size of electrons and atoms). Participants were provided with the scientifically consensual answer to the basic science questions, and asked whether they accept it. Acceptance of the scientific consensus was very high in the sample as a whole (95.1%), but also in every sub-sample (e.g., no trust in science: 87.3%; complete endorsement of flat Earth theory: 87.2%). This quasi-universal acceptance of basic science suggests that people are motivated to reject specific scientific beliefs, and not science as a whole. This could be leveraged in science communication.
keywords          : Trust in science; conspiracy theories; conspiracy thinking; science knowledge.
floatsintext      : yes
linenumbers       : no 
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "doc" # "doc" for nice look, "man" for manuscripty
output: 
  papaja::apa6_docx: default
  papaja::apa6_pdf: 
    keep_tex: true
always_allow_html: true
appendix:
  - "appendix_acceptance_knowledge.Rmd"
  - "appendix_conspiracy_thinking.Rmd"
  - "appendix_trust.Rmd"
  - "appendix_conspiracy_belief.Rmd"
  - "appendix_knowledge.Rmd"
  - "appendix_by-question.Rmd"
  - "appendix_conspiracy-scales.Rmd"
  - "appendix_exp1.Rmd"
  - "appendix_exp2.Rmd"
  - "appendix_exp3.Rmd"
  - "appendix_exp4.Rmd"
bibliography: references.bib
---

>>>>

  Substantial minorities of the population report a low degree of trust in science, or endorse conspiracy theories that violate basic scientific knowledge. This might indicate a wholesale rejection of science. In four studies, we asked 782 US participants questions about trust in science, conspiracy beliefs, and basic science (e.g. the relative size of electrons and atoms). Participants were provided with the scientifically consensual answer to the basic science questions, and asked whether they accept it. Acceptance of the scientific consensus was very high in the sample as a whole (95.1%), but also in every sub-sample (e.g., no trust in science: 87.3%; complete endorsement of flat Earth theory: 87.2%). This quasi-universal acceptance of basic science suggests that people are motivated to reject specific scientific beliefs, and not science as a whole. This could be leveraged in science communication.
keywords          : Trust in science; conspiracy theories; conspiracy thinking; science knowledge.
floatsintext      : yes
linenumbers       : no 
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "doc" # "doc" for nice look, "man" for manuscripty
output: 
  papaja::apa6_pdf: 
    keep_tex: true
  papaja::apa6_docx: default
always_allow_html: true
appendix:
  - "appendix_acceptance_knowledge.Rmd"
  - "appendix_conspiracy_thinking.Rmd"
  - "appendix_trust.Rmd"
  - "appendix_conspiracy_belief.Rmd"
  - "appendix_knowledge.Rmd"
  - "appendix_by-question.Rmd"
  - "appendix_conspiracy-scales.Rmd"
  - "appendix_exp1.Rmd"
  - "appendix_exp2.Rmd"
  - "appendix_exp3.Rmd"
  - "appendix_exp4.Rmd"
bibliography: references.bib
---

### 2025-07-07 paper.tex
second round  revisions update preprint

\abstract{%
Substantial segments of the population report a low degree of trust in science, or endorse conspiracy theories that violate basic scientific knowledge. This might indicate a wholesale rejection of science. In four studies, we asked 782 US participants questions about trust in science, conspiracy beliefs, and basic science (e.g.~the relative size of electrons and atoms). Participants were provided with the scientifically consensual answer to the basic science questions, and asked whether they accept it. Acceptance of the scientific consensus was very high in the sample as a whole (95.1\%), but also in every sub-sample (e.g., no trust in science: 87.3\%; complete endorsement of flat Earth theory: 87.2\%). This quasi-universal acceptance of basic science suggests that people are motivated to reject specific scientific beliefs, and not science as a whole. This could be leveraged in science communication.
}

@@

Because recruitment was conducted through Prolific's system, we do not have access to the number of individuals who viewed the study invitation but chose not to participate. As such, it is not possible for us to calculate response rates (the proportion of invited participants who completed the survey) for the studies. However, we employed attention checks and, for each study, we report exclusions based on these checks.

@@

Why do most people (in the US) trust most of basic science? Based on the present studies, we can only speculate. But if this trust has a rational basis, then education is the most plausible candidate for explaining it. Large-scale surveys have identified education, and in particular science education, as one of the strongest correlates of trust in science (Noy \& O'Brien, 2019; Wellcome Global Monitor, 2018) but not consistently so (see e.g., Cologna et al., 2025 who only find a small positive relationship between tertiary education and trust in science). If these studies were to assess trust in scientific knowledge, rather than assessing trust more broadly, we might expect effects to be more strongly and more coherently positive. Differentiating the impact of education on different facets of trust would be an interesting avenue for future research.

The present studies have a number of limitations, in particular the lack of representative samples, and the focus on a single country. The use of online samples recruited via Prolific Academic does not allow us to calculate response rates and thus estimate non-response bias. However, by targeting vaccine-skeptic samples in some of the studies, we made an effort in recruiting those participants that can be assumed to be most reluctant to partake in a study confronting them with science facts. While our results support motivated reasoning accounts of trust in science, they leave important questions unaddressed, in particular: Why do people reject specific scientific findings but not others? Suggestions have already been made for a number of issues such as vaccination (Miton \& Mercier, 2015), GMOs (Blancke, Van Breusegem, De Jaeger, Braeckman, \& Van Montagu, 2015), or nuclear energy (Hacquin, Altay, Aarøe, \& Mercier, 2021). However, research is still needed to better understand what motivates these rejections (see e.g., Hornsey, 2020). Moreover, rejection of scientific facts can manifest in various ways, for example calling into question the integrity of scientists (Mede, 2023; Mede, Schäfer, Metag, \& Klinger, 2022), or denying the very possibility of scientifically investigating certain issues (Munro, 2010). The classification of justifications we present here does not address these processes in detail. Finally, by design, most of the basic science knowledge presented in this study didn't directly relate to anything controversial for most people. Future studies could look at basic science that does relate to scientific topics which are the object of public and policy-relevant debates (e.g.~GM food, climate science, vaccines), or scientific findings for which there is a more obvious potential conflict of interest (e.g.~non-publicly funded medical research).

>>>>

\abstract{%
Substantial minorities of the population report a low degree of trust in science, or endorse conspiracy theories that violate basic scientific knowledge. This might indicate a wholesale rejection of science. In four studies, we asked 782 US participants questions about trust in science, conspiracy beliefs, and basic science (e.g.~the relative size of electrons and atoms). Participants were provided with the scientifically consensual answer to the basic science questions, and asked whether they accept it. Acceptance of the scientific consensus was very high in the sample as a whole (95.1\%), but also in every sub-sample (e.g., no trust in science: 87.3\%; complete endorsement of flat Earth theory: 87.2\%). This quasi-universal acceptance of basic science suggests that people are motivated to reject specific scientific beliefs, and not science as a whole. This could be leveraged in science communication.
}

@@

Because recruitment was conducted through Prolific, we do not have access to the number of individuals who viewed the study invitation but chose not to participate. As such, it is not possible for us to calculate response rates (the proportion of invited participants who completed the survey) for the studies. However, we employed attention checks and, for each study, we report exclusions based on these checks. Moreover, the study invitation was purposefully nondescript (``short psychology experiment''), so that it would be impossible for participants to self-select because of their attitudes towards science for instance.

@@

Why do most people (in the US) trust most of basic science? Based on the present studies, we can only speculate. But if this trust has a rational basis, then education--and science education in particular--is the most likely explanation. Large-scale surveys have identified science education as one of the strongest, or the strongest, correlates of trust in science (Noy \& O'Brien, 2019; Wellcome Global Monitor, 2018) If these studies were to assess trust in basic scientific knowledge, rather than assessing trust in science more broadly, we might expect effects to be even stronger. Differentiating the impact of education on different facets of trust would be an interesting avenue for future research.

The present studies have a number of limitations, in particular the lack of representative samples, and the focus on a single country. The use of online samples recruited via Prolific does not allow us to calculate response rates and thus estimate non-response bias. However, participants could not have self-selected on the basis of the study topic and, by targeting vaccine-skeptic samples in some of the studies, we made an effort in recruiting those participants that can be assumed to be most reluctant to partake in a study confronting them with science facts. While our results support motivated reasoning accounts of trust in science, they leave important questions unaddressed, in particular: Why do people reject specific scientific findings but not others? Suggestions have already been made for a number of issues such as vaccination (Miton \& Mercier, 2015), GMOs (Blancke, Van Breusegem, De Jaeger, Braeckman, \& Van Montagu, 2015), or nuclear energy (Hacquin, Altay, Aarøe, \& Mercier, 2021). However, research is still needed to better understand what motivates these rejections (see e.g., Hornsey, 2020). Moreover, rejection of scientific facts can manifest in various ways, for example calling into question the integrity of scientists (Mede, 2023; Mede, Schäfer, Metag, \& Klinger, 2022), or denying the very possibility of scientifically investigating certain issues (Munro, 2010). The classification of justifications we present here does not address these processes in detail. Finally, by design, most of the basic science knowledge presented in this study didn't directly relate to anything controversial for most people. Future studies could look at basic science that does relate to scientific topics which are the object of public and policy-relevant debates (e.g.~GM food, climate science, vaccines), or scientific findings for which there is a more obvious potential conflict of interest (e.g.~non-publicly funded medical research).

### 2025-07-07 preprint.Rmd
second round  revisions update preprint

  Many people report a low degree of trust in science, or endorse conspiracy theories that violate basic scientific knowledge. This might indicate a wholesale rejection of science in substantial segments of the population. In four studies, we asked 782 US participants questions about trust in science, conspiracy beliefs, and basic science (e.g. the relative size of electrons and atoms). Participants were provided with the scientifically consensual answer to the basic science questions, and asked whether they accept it. Acceptance of the scientific consensus was very high in the sample as a whole (95.1%), but also in every sub-sample (e.g., no trust in science: 87.3%; complete endorsement of flat Earth theory: 87.2%). This quasi-universal acceptance of basic science suggests that people are motivated to reject specific scientific beliefs, and not science as a whole. This could be leveraged in science communication.
keywords          : Trust in science; conspiracy theories; conspiracy thinking; science knowledge.
floatsintext      : yes
linenumbers       : no 
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "doc" # "doc" for nice look, "man" for manuscripty
output: 
  papaja::apa6_pdf: 
    keep_tex: true
  papaja::apa6_docx: default
always_allow_html: true
appendix:
  - "appendix_acceptance_knowledge.Rmd"
  - "appendix_conspiracy_thinking.Rmd"
  - "appendix_trust.Rmd"
  - "appendix_conspiracy_belief.Rmd"
  - "appendix_knowledge.Rmd"
  - "appendix_by-question.Rmd"
  - "appendix_conspiracy-scales.Rmd"
  - "appendix_exp1.Rmd"
  - "appendix_exp2.Rmd"
  - "appendix_exp3.Rmd"
  - "appendix_exp4.Rmd"
bibliography: references.bib
---

>>>>

  Substantial minorities of the population report a low degree of trust in science, or endorse conspiracy theories that violate basic scientific knowledge. This might indicate a wholesale rejection of science. In four studies, we asked 782 US participants questions about trust in science, conspiracy beliefs, and basic science (e.g. the relative size of electrons and atoms). Participants were provided with the scientifically consensual answer to the basic science questions, and asked whether they accept it. Acceptance of the scientific consensus was very high in the sample as a whole (95.1%), but also in every sub-sample (e.g., no trust in science: 87.3%; complete endorsement of flat Earth theory: 87.2%). This quasi-universal acceptance of basic science suggests that people are motivated to reject specific scientific beliefs, and not science as a whole. This could be leveraged in science communication.
keywords          : Trust in science; conspiracy theories; conspiracy thinking; science knowledge.
floatsintext      : yes
linenumbers       : no 
draft             : no
mask              : no
figurelist        : no
tablelist         : no
footnotelist      : no
classoption       : "doc" # "doc" for nice look, "man" for manuscripty
output: 
  papaja::apa6_pdf: 
    keep_tex: true
  papaja::apa6_docx: default
always_allow_html: true
appendix:
  - "appendix_acceptance_knowledge.Rmd"
  - "appendix_conspiracy_thinking.Rmd"
  - "appendix_trust.Rmd"
  - "appendix_conspiracy_belief.Rmd"
  - "appendix_knowledge.Rmd"
  - "appendix_by-question.Rmd"
  - "appendix_conspiracy-scales.Rmd"
  - "appendix_exp1.Rmd"
  - "appendix_exp2.Rmd"
  - "appendix_exp3.Rmd"
  - "appendix_exp4.Rmd"
bibliography: references.bib
---
